---
title: "Exploratory Analysis of Yeast Image Data"
author: Austin Chang, Luke Checchio, Daniel Cohen
date: Jan 28, 2025
output: 
  beamer_presentation:
    incremental: false  
    theme: "Frankfurt"  
    colortheme: "rose"  
    toc: true   
    slide_level: 3
    keep_tex: true
header-includes: 
- \AtBeginSubsection{}
---

# Data Grouping/Clustering
##

### Process

- Building on exploratory work by Dr. Iversen
- Wanted to explore whether there was clustering amongst images across first 6 principal components (explained later)
- Extracted PC scores for the first six PCs for each observation
- Calculated median PC scores for the first six PCs for each image
- Used `hclust()` with complete linkage to visualize clustering
- Used `heatmap.2()` to further visualize clustering amongst the images and PCs

### Results

```{r}
#| label: PCA-clusters
#| echo: FALSE
#| warning: FALSE
#| message: FALSE

library(tidyverse)
library(knitr) 
library(ggfortify)
library(gplots)

load("./data/yeast.RData")
data_scaled <- data.frame(scale(y[, sapply(y, is.numeric)])) |> select(!c(Location_Center_X,
                                                              Location_Center_Y))

pca <- prcomp(data_scaled, center = TRUE, scale. = TRUE)

pc_scores <- pca$x[, 1:6]
pc_scores_new <- data.frame(pc_scores) |> rownames_to_column(var = "rowname") |>
  mutate(img = substring(rowname, 0, 8)) |>
  select(!rowname)

pc_scores_sum <- pc_scores_new |> group_by(img) |>
  summarize(PC1.medn = median(PC1), PC2.medn = median(PC2), PC3.medn = median(PC3),
            PC4.medn = median(PC4), PC5.medn = median(PC5), PC6.medn = median(PC6)) |>
  column_to_rownames(var = "img")
dist_matrix <- dist(pc_scores_sum)


hc_result <- hclust(dist_matrix, method = "complete")
plot(hc_result, main = "Hierarchical Clustering on First 6 PCs", xlab = "", sub = "")
```

### Results

```{r}
#| label: PCA-heatmap
#| echo: FALSE
#| warning: FALSE
#| message: FALSE

heatmap.2(as.matrix(pc_scores_sum),scale="none",trace="none",
          cexCol=1)
```

# Feature Selection/Specification
## 

### Principal Component Analysis

```{r}
#| label: PCA-importance
#| echo: FALSE
#| warning: FALSE
#| message: FALSE

sum <- summary(pca)
data.frame(sum$importance) |> select(PC1, PC2, PC3, PC4, PC5, PC6) |> kable(digits = 3)
```


- Include Skree plot or kable output from `prcomp()`
- Explain decision to include first six PCs (>95% cum. variance explained)
- Explain weighted loading metric to determine importance across first six PCs
- Identify important features based on this

# Outlier Identification

- Want to mention potential use of Mahalanobis Distance
- For now, using Euclidean distance

# Data Visualization

- Add plots visualizing important features relationships with strain/well/image

